---
title: "Comparing Distributions"
author:
- name: Kieran Healy
  url: https://kieranhealy.org
  affiliation: Duke University
  affiliation_url: https://sociology.duke.edu
date: "`r Sys.Date()`"
description: |
  Working document.
toc: false  
abstract: Working document only
output:
  html_document:
    highlight: tango    
  pdf_document:
    template: ~/.pandoc/templates/rmd-latex.template  
---

```{r, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 300)

library(here)
library(tidyverse)
library(ggridges)

## Good cb-friendly palette
library(ggokabeito)

## Fonts aren't necessary to reproduce.
## Comment out and change the default theme as needed.
library(myriad)
library(showtext)
showtext_auto()
import_myriad_semi()

theme_figs <- function(){
  theme_myriad_semi() +
    theme(
      plot.background = element_rect(color = "white"),
      plot.title = element_text(size = rel(1.5)),
      plot.subtitle = element_text(size = rel(1.2)),
      axis.title.x = element_text(size = rel(1.25)),
      axis.title.y = element_text(size = rel(1.25)),
#      axis.text.x = element_text(size = rel(1.2)),
      axis.text.y.left = element_text(size = rel(1.2)),
      axis.text.y.right = element_text(size = rel(0.9)))
}

# theme_set(theme_minimal())
theme_set(theme_figs())

## Colors for manual use as needed
my_oka <- palette_okabe_ito(order = c(1, 2, 3, 5, 7), 
                            alpha = NULL)
```

```{r}
## Keep track of labels for as_labeller() functions in faceting.
grp_names <- c(`a` = "Group A",
               `b` = "Group B",
               `c` = "Group C",
               `pop_a` = "Group A",
               `pop_b` = "Group B",  
               `pop_c` = "Group C",  
               `pop_total` = "Total",                 
               `A` = "Group A", 
               `B` = "Group B", 
               `C` = "Group C")

```


## Generate some data

Say we have some number of observed units, e.g., three thousand "counties" or whatever. Each county has some population. Across all counties the population is distributed log-normally around some mean county population. Populations are divided into groups within counties whose proportions vary across counties but always sum to one within each county. 

```{r}
set.seed(1243098)

N <- 3e3
alphas <- c(1.5, 0.9, 2)

pop <- round(rlnorm(N, meanlog = 10.3, sdlog = 1.49), 0)

df <- as_tibble(gtools::rdirichlet(N, alphas), 
                       .name_repair = "unique") %>% 
  rename_with(~ c("a", "b", "c")) %>% 
  rowid_to_column("unit") %>% 
  add_column(pop_total = pop) %>% 
  mutate(across(a:c, 
                .fns = list(pop = ~ round(.x * pop_total, 0) + 1),
                .names = "{fn}_{col}"))

df
```

We can look at the distribution of group populations across counties:

```{r, fig.height=3, fig.width=8, out.width="100%"}

df %>% 
  pivot_longer(a:c) %>% 
  ggplot() + 
  geom_area(mapping = aes(x = value, y = ..count.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5) + 
  scale_fill_manual(values = alpha(my_oka, 0.7)) + 
  scale_color_manual(values = alpha(my_oka, 1)) + 
  guides(color = "none", 
         fill = "none") + 
  labs(x = "Logged Population", y = "Count", 
       title = "Subgroup distribution across units") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 

```



From now on let's just work with the population counts.

```{r}
df <- df %>% 
  select(unit, pop_a:pop_c, pop_total)

df
```

```{r, fig.height=2, fig.width=8, out.width="100%"}
df %>%
  pivot_longer(pop_a:pop_total) %>% 
  group_by(name) %>% 
  summarize(total = sum(value)) %>% 
  ggplot(mapping = aes(x = total, y = name, fill = name)) +  
  geom_col() + 
  guides(fill = "none") + 
  scale_fill_manual(values = alpha(c( my_oka[1:3], "gray40"), 0.9))  + 
  scale_x_continuous(labels = scales::label_number_si()) + 
  scale_y_discrete(labels = as_labeller(grp_names)) +
  labs(y = NULL, x = "Population")
```


Now we can plot the group-level population distributions across counties. We want to compare group distributions to one another and to the overall population distribution by county. 

## Single-panel

```{r, fig.height=5, fig.width=8, out.width="100%"}

df %>%
  pivot_longer(cols = pop_a:pop_total) %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = log(value), y = ..count.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5,
            alpha = 0.7) + 
  scale_color_manual(values = alpha(c( my_oka[1:3], "gray40"), 1),
                     labels = as_labeller(grp_names)) + 
  scale_fill_manual(values = alpha(c( my_oka[1:3], "gray40"), 0.6),
                    labels = as_labeller(grp_names)) + 
  labs(x = "Logged Population", y = "Count", color = "Group", 
       fill = "Group",
       title = "Comparing Subgroups: Histograms", 
       subtitle = "Overall distribution shown in gray")


```


```{r, fig.height=4, fig.width=8, out.width="100%"}
df %>%
  pivot_longer(cols = pop_a:pop_total) %>%
  ggplot() + 
  geom_density(mapping = aes(x = log(value), 
                          color = name, fill = name), 
            alpha = 0.5) + 
  scale_color_manual(values = alpha(c( my_oka[1:3], "gray40"), 1),
                     labels = as_labeller(grp_names)) + 
  scale_fill_manual(values = alpha(c( my_oka[1:3], "gray40"), 0.6),
                    labels = as_labeller(grp_names)) + 
  labs(x = "Logged Population", y = "Density", 
       title = "Comparing Subgroups: Density", 
       color = "Group", 
       fill = "Group")


```


## Ridgelines

```{r, fig.height=5, fig.width=8, out.width="100%"}
df %>%
  pivot_longer(cols = pop_a:pop_total) %>%
  ggplot() + 
  geom_density_ridges(mapping = aes(x = log(value + 1), 
                                    y = name, 
                                    fill = name), 
                      color = "white") + 
  scale_fill_manual(values = alpha(c( my_oka[1:3], "gray40"), 0.7)) + 
  scale_y_discrete(labels = as_labeller(grp_names)) + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = NULL, title = "Comparing Total and Subgroups: Ridgelines") + 
  theme_ridges(font_family = "Myriad Pro SemiCondensed") 



```


## Histograms

Now we're going to facet our plots so that every subgroup gets its own panel. But we want to compare it to the overall distribution. So we pivot differently. The trick is to hold on to the total population by only pivoting the subgroups to long format. That leaves us with repeated values for the total population, `pop_total`, like this: 

```{r}
df %>%
  pivot_longer(cols = pop_a:pop_c)
```

When we draw the plot, we first call on `geom_histogram()` to draw the distribution of the total population, setting the color to gray. Then we call it again, separately, to draw the subgroups. Finally we facet on the subgroup names. This leaves us with a faceted plot where each panel shows one subpopulation's distribution and, for reference behind it, the overall population distribution.

```{r, fig.height=3, fig.width=8, out.width="100%"}
df %>%
  pivot_longer(cols = pop_a:pop_c) %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = log(pop_total), y = ..count..), 
                bins = 20, alpha = 0.7,
                fill = "gray40", size = 0.5) + 
  geom_histogram(mapping = aes(x = log(value), y = ..count.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5,
            alpha = 0.7) + 
  scale_fill_okabe_ito() + 
  scale_color_okabe_ito() + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = "Count", 
       title = "Comparing Subgroups: Histograms", 
       subtitle = "Overall distribution shown in gray") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 
```

This is a handy trick. We'll use it repeatedly in the remaining figures, as we look at different ways of drawing the same comparison. 

```{r, fig.height=3, fig.width=8, out.width="100%"}
df %>% 
  pivot_longer(cols = pop_a:pop_c) %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = log(pop_total), y = ..ncount..), 
                bins = 20, alpha = 0.8,
                fill = "gray40", size = 0.5) + 
  geom_histogram(mapping = aes(x = log(value), y = ..ncount.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5,
            alpha = 0.7) + 
  scale_fill_manual(values = alpha(my_oka, 0.7)) + 
  scale_color_manual(values = alpha(my_oka, 1)) + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = "Scaled Count", 
       title = "Comparing Subgroups: Histograms (Scaled Counts)", 
       subtitle = "Overall distribution shown in gray") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 



```

```{r, fig.height=3, fig.width=8, out.width="100%"}
df %>% 
  pivot_longer(cols = pop_a:pop_c) %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = log(pop_total), y = ..density..), 
                bins = 20, alpha = 0.7,
                fill = "gray40", size = 0.5) + 
  geom_histogram(mapping = aes(x = log(value), y = ..density.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5,
            alpha = 0.7) + 
  scale_fill_manual(values = alpha(my_oka, 0.7)) + 
  scale_color_manual(values = alpha(my_oka, 1)) + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = "Density", 
       title = "Comparing Subgroups: Histograms (Density)", 
       subtitle = "Overall distribution shown in gray") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 



```


```{r, fig.height=3, fig.width=8, out.width="100%"}
df %>% 
  pivot_longer(cols = pop_a:pop_c) %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = log(value), y = ..density.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5,
            alpha = 0.7) + 
  geom_step(mapping = aes(x = log(pop_total), y = ..density..), 
                bins = 20, alpha = 0.9,
                color = "gray30", size = 0.6, 
            stat = "bin",
            direction = "mid") + 
  scale_fill_manual(values = alpha(my_oka, 0.8)) + 
  scale_color_manual(values = alpha(my_oka, 1)) + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = "Density", 
       title = "Comparing Subgroups: Histograms (Density)", 
       subtitle = "Overall distribution shown in outline") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 



```


## Frequency Polygons

```{r, fig.height=3, fig.width=8, out.width="100%"}
df %>% 
  pivot_longer(cols = pop_a:pop_c) %>%
  ggplot() + 
  geom_area(mapping = aes(x = log(value), y = ..count.., 
                          color = name, fill = name), 
            stat = "bin", bins = 20, size = 0.5) + 
  geom_freqpoly(mapping = aes(x = log(pop_total), y = ..count..), 
                bins = 20, 
                color = "gray20", size = 0.5) + 
  scale_fill_manual(values = alpha(my_oka, 0.7)) + 
  scale_color_manual(values = alpha(my_oka, 1)) + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = "Count", 
       title = "Comparing Subgroups: Frequency Polygons", 
       subtitle = "Overall distribution shown in outline") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 



```

## Kernel Densities

```{r, fig.height=3, fig.width=8, out.width="100%"}
df %>% 
  pivot_longer(cols = pop_a:pop_c) %>%
  ggplot() + 
  geom_density(mapping = aes(x = log(value), 
                          color = name, fill = name), 
            size = 0.5) + 
  geom_density(mapping = aes(x = log(pop_total)), 
                color = "gray20", size = 0.5) + 
  scale_fill_manual(values = alpha(my_oka, 0.7)) + 
  scale_color_manual(values = alpha(my_oka, 1)) + 
  guides(color = "none", fill = "none") + 
  labs(x = "Logged Population", y = "Density", 
       title = "Comparing Subgroups: Kernel Densities", 
       subtitle = "Overall distribution shown in outline") + 
  facet_wrap(~ name, nrow = 1, labeller = as_labeller(grp_names)) 



```




```{r, fig.width=12, fig.height=3.5, out.width="100%"}
library(gapminder)
p <- ggplot(data = gapminder, mapping = aes(x = year, y = gdpPercap))
p + geom_line(color="gray70", aes(group = country)) +
    geom_smooth(size = 1.1, method = "loess", se = FALSE) +
    scale_y_log10(labels=scales::dollar) +
    facet_wrap(~ continent, ncol = 5) +
    labs(x = "Year",
         y = "GDP per capita",
         title = "GDP per capita on Five Continents", 
         subtitle = "Individual countries shown in gray, trend in blue.")

```









